### 2020-10-27 doc/manuscript/Dow_manuscript.Rmd
bit of work on intro & cleanup

**Title:** Effect of spring temperatures on tree growth phenology in a temperate deciduous forest

@@

#figures
schematic_caption <- "Figure 1. Schematic illustrating the parameters considered here. Shown are measurements for an example tree (## cm SPECIES), fit with the model of McMahon & Parker (2015), from which phenology and growth rate parameters are obtained. ... (HERE, WE NEED TO ANNOTATE WITH LABELS FOR EACH PARAMETER CONSIDERED IN THE MANUSCRIPT. I'D ALSO DROP THE BACKGROUND SHADING--MAKES IT BUSY.)"

@@

climwin_caption <- "Figure 3. Landscapes of relationships between the day of year on which 25% of annual growth is achieved (DOY_25) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest. Shown are matrices of linear coefficients of first-order linear regressions between temperature and DOY_25, where Window Open and Window Close indicate number of weeks prior to DOY_25 (ring-porous: DATE at SCBI, DATE at HF; diffuse-porous: DATE at SCBI, DATE at HF). Black boxes indicate the optimal window (i.e., that with lowest dAIC).    (REPLOT ON THE SAME SCALE, WITH BLUE FOR POSITIVE AND RED FOR NEGATIVE)"

@@

#### (2. spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
In recent decades, an increase in average Spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006).** The effects of this Spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations*. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of this Spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).

#### (3. However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). However, the fine-scale responses of temperate deciduous tree growth to a warming Spring are not commonly recorded due to the increased effort required (). Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).

@@

Earlier leaf up timings could signal that these trees are shifting their growth windows in response to increasing Spring temperatures, but to date, very little research has been done on topics like these. In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in Spring temperature patterns. Warming events in late-Spring were shown to shift leaf emergence timing back more than early-Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016). Identifying the pre-season period of greatest sensitivity of intraannual growth, along with the effect warming within that period has on other stem phenology traits could help unravel the mysteries of tree growth.

@@

The significance of wood type in regards to wood and leaf phenology likely stems from evolutionary differences and adaptations to local climates where these trees can be found. Ring-porous trees have large xylem cells which are prone to embolism during the winter. These cells are repaired before leaf development, explaining why the 25% yearly growth milestone is reached around the same time as bud burst. Diffuse-porous xylem cells are smaller, and less prone to embolism during the Winter, meaning they do not need to be repaired in order to restore hydraulic functionality before beginning the leaf development process. Results from the paper by D’Orangeville et al. (in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during Spring, while diffuse-porous trees reach these milestones in the Summer, but both reached 75% total annual growth around the same time. If leaf development is sped up when Spring temperatures are warmer, we should seek to understand whether this relationship holds, and how it potentially varies among trees of different wood-types.

@@

The period where the phenology milestone(s) were most affected by changes in early Spring temperature was determined using the R package Climwin. This package tests the correlation between climate variables (TMAX) and biological variables (DOY, max rate DOY, total growth) within a specified time-frame, reporting correlation and delta AIC's of each 'window'. as the best-model. We instructed Climwin to search for the best window beginning at February 1st and ending at the mean DOY for each phenology variable (chart in SI?). This was done to find corresponding (median of 95% CI window open and close) start and end dates for the analysis of ring-porous and diffuse-porous trees, as well as all trees pooled together.

@@

Correlation between these variables and Spring temperatures over the 15 year study period (1998-2003; 2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.

@@

A mixed effect model was used to test the response of wood phenology variables (25% DOY, 50% DOY, 75% DOY, max-rate, max-rate DOY, total growth) to fixed effects of wood-type and Spring temperature, along with random effects of species and tag. We ran two separate models of each wood phenology variable for each of our major wood-types, and one combined model to use for comparison. Models were run using the LME4 package in R (Bates, D., Mächler, M., Bolker, B. & Walker, S. (2015). Fitting linear mixed‐effects models using lme4. J. Stat. Softw., 67, 1–48.)

@@

We have shown that intraannual growth patterns in North American temperate trees change in response to increasing Spring temperatures. The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here. Ring-porous trees must repair winter embolism of their large xylem vessels before leaf expansion, achieving roughly 25% total yearly growth simultaneously with bud burst, and 50% yearly growth around the time of 75% leaf emergence (D'Orangeville et al,). On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al). These differences are likely evolutionary hydraulic adaptations to climate (Baas & Wheeler, 2011). The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the Spring at the cost of increased cavitation risk during Spring drought, and high likeliehood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994). The pre-leaf stem growth occurring in these species is fueled by non-structural carbon stores, often used for other cell maintenance activities within the plant (Sperry et al, 1994; Palacio et al, 2014; Sevanto et al, 2014). The dependence on these carbon stores for cambial activation could prove to be detrimental to long-term health of the tree, as exhaustion of reserves leads to carbon starvation (Sevanto et al, 2014). Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the Spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer (Sperry et al, 1994). This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.

@@

The mechanisms controlling activation of cambium reactivation is still debated, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves (Frankenstein et al, 2005). However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production. *READ MORE ABOUT THIS!* The relationship between earlier cambial reactivation and early Spring temperature signals that the mechanism is influenced by temperature?

Under the warming regime analyzed here, ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while timing of 75% growth !!! Changed by constant amount !!! . This implies that the growing season of ring-porous trees expanded due to Spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies. In fact, we found a positive correlation between ring-porous growing season length and Spring temperature, while no such relationship was observed in diffuse-porous trees. The earlier reactivation of cambial cells is potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990). Damage of this nature would then need to be repaired using carbon reserves, which may be in short demand as they are already being used for initial growth.This would negatively impact the plant's ability to maintain cell health, or reduce the ability to respond to invasion of pests or pathogens. Alternatively, the extension of growth season and higher yearly growth totals like we see here, could give ring-porous trees a competitive advantage over their non-responsive diffuse-porous counterparts.

Diffuse-porous growth timing seems relatively unaffected by early Spring temperatures, instead showing a higher correlation with late Spring or Summer temps. This disconnect means that diffuse-porous trees will continue to focus their growing season in the Summer, which is projected to become less conducive to tree growth as temperatures and drought intensity increase in some areas. ---- Studies have found that diffuse-porous trees in the Eastern USA show a larger reduction in ring-width increment during, and following Summer droughts when compared to species of other wood types (Orwig & Abrams, 1997; Kannenberg et al, 2018) as well as greater sensitivity to Summer precipitation patterns (Elliot et al, 2015). The potential for more frequent reduction in growth coupled with the increased growing season length of ring-porous species may lead to an alteration in forest species composition as diffuse-porous species are outcompeted.

@@

This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is only the tip of the ice berg. Here, we focused our analysis purely on early Spring warming. This falls short of the reality of our changing climate, where each season is being affected in some way, leading to a combination of stressors on plant growth. For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June TMAX on trees of that wood type (figure or table?). This revelation agrees with the results from other climate sensitivity studies like Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year. Additionally, Other studies have found that Autumn leaf senescence is being delayed in response to warmer temperatures in the previous year's Winter or Autumn (), positive correlation between growth and higher winter temperatures (Heide, 1993; Delpierre et al, 2009), stunted total productivity induced by drought and high TMAX in the current year's Summer (), and an increase in diffuse-porous tree growth during the growing season following an abnormally cold winter (Benito & Pederson, 2015; Elliot et al, 2015). How these relationships may impact intraannual growth patterns is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.

@@

Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the Spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).

>>>>

**Title:** Effect of spring temperatures on tree growth phenology in two temperate deciduous forests

@@

#figures
schematic_caption <- "Figure 1. Schematic illustrating the parameters considered here. Shown are measurements for an example tree (## cm SPECIES), fit with the model of McMahon & Parker (2015), from which phenology and growth rate parameters are obtained."

@@

climwin_caption <- "Figure 3. Landscapes of relationships between the day of year on which 25% of annual growth is achieved (DOY_25) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest. Shown are matrices of linear coefficients of first-order linear regressions between temperature and DOY_25, where Window Open and Window Close indicate number of weeks prior to DOY_25 (ring-porous: DATE at SCBI, DATE at HF; diffuse-porous: DATE at SCBI, DATE at HF). Black boxes indicate the optimal window."

@@

#### (2. spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
In recent decades, an increase in average spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). 
Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). 
The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations*. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of this spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).

Many observations of climate induced phenological shifts are based on GIS (), eddy-covarience measurements (), or other broad scale measurements of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review).

#### (3. However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )

Earlier leaf up timings could signal that these trees are shifting their growth windows in response to increasing spring temperatures, but to date, very little research has been done on this topic. 
In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in spring temperature patterns. *(but check into this.... I suspect there may be some that we're missing)*
Warming events in late-spring were shown to shift leaf emergence timing back more than early-spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016). Identifying the pre-season period of greatest sensitivity of intraannual growth, along with the effect warming within that period has on other stem phenology traits could help unravel the mysteries of tree growth.

@@

@@

The significance of wood type in regards to wood and leaf phenology likely stems from evolutionary differences and adaptations to local climates where these trees can be found. Ring-porous trees have large xylem cells which are prone to embolism during the winter. These cells are repaired before leaf development, explaining why the 25% yearly growth milestone is reached around the same time as bud burst. Diffuse-porous xylem cells are smaller, and less prone to embolism during the Winter, meaning they do not need to be repaired in order to restore hydraulic functionality before beginning the leaf development process. Results from the paper by D’Orangeville et al. (in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during spring, while diffuse-porous trees reach these milestones in the Summer, but both reached 75% total annual growth around the same time. If leaf development is sped up when spring temperatures are warmer, we should seek to understand whether this relationship holds, and how it potentially varies among trees of different wood-types.

@@

The period where the phenology milestone(s) were most affected by changes in early spring temperature was determined using the R package Climwin. This package tests the correlation between climate variables (TMAX) and biological variables (DOY, max rate DOY, total growth) within a specified time-frame, reporting correlation and delta AIC's of each 'window'. as the best-model. We instructed Climwin to search for the best window beginning at February 1st and ending at the mean DOY for each phenology variable (chart in SI?). This was done to find corresponding (median of 95% CI window open and close) start and end dates for the analysis of ring-porous and diffuse-porous trees, as well as all trees pooled together.

@@

Correlation between these variables and spring temperatures over the 15 year study period (1998-2003; 2011-2019) were assessed using a linear mixed model in a heriarchal bayesian framework.

@@

A mixed effect model was used to test the response of wood phenology variables (25% DOY, 50% DOY, 75% DOY, max-rate, max-rate DOY, total growth) to fixed effects of wood-type and spring temperature, along with random effects of species and tag. We ran two separate models of each wood phenology variable for each of our major wood-types, and one combined model to use for comparison. Models were run using the LME4 package in R (Bates, D., Mächler, M., Bolker, B. & Walker, S. (2015). Fitting linear mixed‐effects models using lme4. J. Stat. Softw., 67, 1–48.)

@@

We have shown that intraannual growth patterns in North American temperate trees change in response to increasing spring temperatures. The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here. Ring-porous trees must repair winter embolism of their large xylem vessels before leaf expansion, achieving roughly 25% total yearly growth simultaneously with bud burst, and 50% yearly growth around the time of 75% leaf emergence (D'Orangeville et al,). On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al). These differences are likely evolutionary hydraulic adaptations to climate (Baas & Wheeler, 2011). The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the spring at the cost of increased cavitation risk during spring drought, and high likeliehood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994). The pre-leaf stem growth occurring in these species is fueled by non-structural carbon stores, often used for other cell maintenance activities within the plant (Sperry et al, 1994; Palacio et al, 2014; Sevanto et al, 2014). The dependence on these carbon stores for cambial activation could prove to be detrimental to long-term health of the tree, as exhaustion of reserves leads to carbon starvation (Sevanto et al, 2014). Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer (Sperry et al, 1994). This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.

@@

The mechanisms controlling activation of cambium reactivation is still debated, as some speculate Auxin, the chemical believed to kick-start the reactivation process, originates in the leaves (Frankenstein et al, 2005). However, the fact that ring-porous species complete 25% growth before the presence of leaf buds indicates there may be another source of Auxin production. *READ MORE ABOUT THIS!* The relationship between earlier cambial reactivation and early spring temperature signals that the mechanism is influenced by temperature?

Under the warming regime analyzed here, ring-porous trees reach 25% and 50% growth milestones earlier (insert actual # days?) in the year, while timing of 75% growth !!! Changed by constant amount !!! . This implies that the growing season of ring-porous trees expanded due to spring temperatures, which are projected to continue to increase as anthropogenic global warming intensifies. In fact, we found a positive correlation between ring-porous growing season length and spring temperature, while no such relationship was observed in diffuse-porous trees. The earlier reactivation of cambial cells is potentially exposing these trees to higher risk of cavitation or damage during frost events (Cochard & Tyree, 1990). Damage of this nature would then need to be repaired using carbon reserves, which may be in short demand as they are already being used for initial growth.This would negatively impact the plant's ability to maintain cell health, or reduce the ability to respond to invasion of pests or pathogens. Alternatively, the extension of growth season and higher yearly growth totals like we see here, could give ring-porous trees a competitive advantage over their non-responsive diffuse-porous counterparts.

Diffuse-porous growth timing seems relatively unaffected by early spring temperatures, instead showing a higher correlation with late spring or Summer temps. This disconnect means that diffuse-porous trees will continue to focus their growing season in the Summer, which is projected to become less conducive to tree growth as temperatures and drought intensity increase in some areas. ---- Studies have found that diffuse-porous trees in the Eastern USA show a larger reduction in ring-width increment during, and following Summer droughts when compared to species of other wood types (Orwig & Abrams, 1997; Kannenberg et al, 2018) as well as greater sensitivity to Summer precipitation patterns (Elliot et al, 2015). The potential for more frequent reduction in growth coupled with the increased growing season length of ring-porous species may lead to an alteration in forest species composition as diffuse-porous species are outcompeted.

@@

This analysis is one of a first generation of intraannual dendroband tree growth studies, and as such, is only the tip of the ice berg. Here, we focused our analysis purely on early spring warming. This falls short of the reality of our changing climate, where each season is being affected in some way, leading to a combination of stressors on plant growth. For example, diffuse porous trees seemed to show little, if any, relationship with the climate window defined here, but if we shift that window forward 20 days, the effect increases dramatically, signalling a greater importance of May/June TMAX on trees of that wood type (figure or table?). This revelation agrees with the results from other climate sensitivity studies like Helcoski et al (2019), which found differences in interspecific growth responses to several climate variables throughout the year. Additionally, Other studies have found that Autumn leaf senescence is being delayed in response to warmer temperatures in the previous year's Winter or Autumn (), positive correlation between growth and higher winter temperatures (Heide, 1993; Delpierre et al, 2009), stunted total productivity induced by drought and high TMAX in the current year's Summer (), and an increase in diffuse-porous tree growth during the growing season following an abnormally cold winter (Benito & Pederson, 2015; Elliot et al, 2015). How these relationships may impact intraannual growth patterns is uncertain, but identifying such relationships will help us further understand how climate change is impacting forests.

@@

Another study found that the velocity of leaf development is correlated to bud burst timing, such that the later in the spring bud burst occurred, the faster leaves developed and vice versa (Klosterman & Hufkins & Richardson, 2018).
